### 2022-03-25 ehoh2/ehoh2.tex
unif

>>>>

\section{Unification, Matching, and Term Indexing}
\label{sec:ehoh2:unif-match-index}

Standard superposition crucially depends on the concept of a most general
unifier (MGU). In higher-order logic, such a unifier does not in general exist,
and the concept
%%% @PETAR: CSU, infinite
is replaced by that of a complete set of unifiers (CSU), which may be infinite.
In Chapter \ref{ch:unif}, we described an efficient procedure to enumerate a CSU
for a term pair. It is implemented in Zipperposition, together with some
extensions to term indexing. In \ehohii{}, we further improve the performance of
this procedure by implementing a terminating, incomplete variant. We also
introduce a new indexing data structure.

\subsection{The Unification Procedure} 
% Two main guiding principles behind our
% unification procedure are laziness and redundancy elimination. The first one is
% embodied in the fact that the procedure makes sure that it never fully
% normalizes the terms or fully applies the substitution if that is not necessary.
% Redundancy elimination is achieved through \emph{oracles}
% \cite{unif-section}: the procedures that solve the unification
% problem for a subclass of higher-order terms on which unification is decidable,
% and for \ehohii{}, unary. If those procedures determine that the given terms
% belong to their class of terms, they will return the most general unifier,
% avoiding redundant computations and in some cases even non-termination.
The unification procedure works by maintaining a list of unification pairs to be solved.
After choosing a pair, it first normalizes it by $\beta$-reducing and
instantiating the heads of both terms in the pair. Then, if either head is a
variable, it computes an appropriate binding for this variable, thereby
approximating the solution.

Unlike in first-order and $\lambda$-free higher-order unification, in full
higher-order case there may be many bindings that lead to a solution. To reduce
this mostly blind guessing of bindings, the procedure features support for
\emph{oracles} (Sect.~\ref{sec:unif:the-unification-procedure}). These are
procedures that solve the unification problem for a subclass of higher-order
terms on which unification is decidable and, in the case of \ehohii{}, unary. Oracles help
increase performance, avoid nontermination, and avoid redundant bindings.

\looseness=-1
In Chapter \ref{ch:unif}, the unification procedure is described as a transition system. In
\ehohii{}, the procedure is implemented nonrecursively, and the unifiers are
enumerated using an iterator object that encapsulates the unifier
search state. The iterator consists of five fields:
\begin{enumerate}
  \item \vn{constraints}, which holds the unification
  constraints
  \item \vn{bt\_state}, a stack that contains information necessary to backtrack
  to a previous state
  \item \vn{branch\_iter}, which stores how far we
  are in exploring different possibilities from the current search node
  \item \vn{steps}, which remembers how many different unification bindings (such as
  imitation, projection, and identification) are applied
  \item \vn{subst},
  a stack storing the variables bound so far
\end{enumerate}
\pagebreak[2]
\newcommand{\cn}[1]{\ensuremath{\textsc{#1}}} %const name
\newcommand{\fc}[1]{\ensuremath{\textsc{#1}}} %function call
\algrenewcommand\alglinenumber[1]{{\color{gray} \tiny #1}}

The iterator is initialized to hold the original problem in \vn{constraints},
and all other fields are initially empty. The unifiers are retrieved one by one by
calling the function \fc{ForwardIter}. It returns \fc{True} if the iterator made
progress, in which case the unifier can be read via the iterator's
$\vn{subst}$ field. Otherwise, no more unifiers can be found, and the iterator is
no longer valid. The function's pseudocode is given below, including two
auxiliary functions \fc{NormalizeHead} and \fc{BacktrackIter}:

\vspace{\jot}
\begin{algorithmic}[]
  \Function{ForwardIter}{$\vn{iter}$}
  \State $\vn{forward} \gets \neg \vn{iter.constraints.empty()} \lor \fc{BacktrackIter}(\vn{iter})$
  \While{$\vn{forward} \land \neg \vn{iter.constraints.empty()}$}
    \State $(\vn{lhs}, \vn{rhs}) \gets $ pop pair from $\vn{iter.constraints}$
    \State $\vn{lhs} \gets \fc{NormalizeHead}(\vn{lhs})$
    \State $\vn{rhs} \gets \fc{NormalizeHead}(\vn{rhs})$
    \State normalize and discard the $\lambda$ prefix of $\vn{lhs}$ and $\vn{rhs}$

    \If{$\neg\vn{lhs.head.is\_var()} \land \vn{rhs.head.is\_var()}$}
      \State swap $\vn{lhs}$ and $\vn{rhs}$
    \EndIf

    \If{$\vn{lhs.head.is\_var()}$}
      \State $\vn{oracle\_res} \gets \fc{Fixpoint}(\vn{lhs, rhs, iter.subst}) $
      \If{$\vn{oracle\_res} = \cn{NotInFragment}$}
        \State $\vn{oracle\_res} \gets \fc{Pattern}(\vn{lhs, rhs, iter.subst}) $
      \EndIf

      \If{$\vn{oracle\_res} = \cn{NotUnifiable}$}
        \State $\vn{forward} \gets \fc{BacktrackIter(iter)}$
      \ElsIf{$\vn{oracle\_res} = \cn{NotInFragment}$}
        \State $\vn{n\_steps}, \vn{n\_branch\_iter}, \vn{n\_binding} \gets$
        \State $\quad \fc{NextBinding}(\vn{lhs}, \vn{rhs}, \vn{iter.steps}, \vn{iter.branch\_iter})$

        \If{$\vn{n\_branch\_iter} \not= \fc{BindEnd}$ }
          \State push pair $(\vn{lhs},\vn{rhs})$ back to \vn{iter.constraints}
          \State push quadruple $(\vn{iter.constraints}, \vn{n\_branch\_iter}, $
          \State                  $\quad \vn{iter.steps}, \vn{iter.subst})$ onto $\vn{iter.bt\_state}$
          \State extend $\vn{iter.subst}$ with $\vn{n\_binding}$
          \State $\vn{iter.steps} \gets \vn{n\_steps}$
          \State $\vn{iter.branch\_iter} \gets \cn{BindBegin}$
        \ElsIf{\vn{lhs.head = rhs.head}}
          \State create constraint pairs of arguments of $\vn{lhs}$ and $\vn{rhs}$
          \State $\quad$ and push them to \vn{iter.constrants}
          \State $\vn{iter.branch\_iter} \gets \cn{BindBegin}$
        \EndIf
      \EndIf
    \ElsIf{$\vn{lhs.head} = \vn{rhs.head}$}
      \State create constraint pairs of arguments of $\vn{lhs}$ and $\vn{rhs}$
      \State $\quad$ and push them to \vn{iter.constrants}
    \Else{} \State $\vn{forward} \gets \fc{BacktrackIter(\vn{iter})}$
    \EndIf
  \EndWhile
  \State \Return $\vn{forward}$
  \EndFunction

  \Function{NormalizeHead}{$\vn{t}$}
    \If{$\vn{t.head} = \internalat \land \vn{t.args}[0].\vn{is\_lambda}()$}
      \State reduce the top-level $\beta$-redex in $t$
      \State \Return $\fc{NormalizeHead}(t)$
    \ElsIf{$\vn{t.head.is\_var()} \land \vn{t.head.binding} \not= \cn{nil}$}
      \State $\vn{t.head} \gets \vn{t.head.binding}$
      \State \Return $\fc{NormalizeHead}(t)$
    \Else{}
      \State \Return $t$
    \EndIf
  \EndFunction

  \vspace{\jot}
  \Function{BacktrackIter}{$\vn{iter}$}
    \If{\vn{iter.bt\_state.empty()}}
      \State clear all fields in \vn{iter}
      \State \Return \cn{False}
    \Else
    \State pop $(\vn{constraints}, \vn{branch\_iter}, \vn{steps}, \vn{subst})$ from $\vn{iter.bt\_state}$
    \State set the corresponding fields of $\vn{iter}$
    \State \Return \cn{True}
    \EndIf
  \EndFunction
  \vspace{\jot}

\end{algorithmic}

\looseness=-1
\fc{ForwardIter} begins by backtracking if the previous attempt was successful (i.e., 
all constraints were solved). If it finds a state from which it can continue,
it takes term pairs from $\vn{constraints}$ until
there are no more constraints or it is determined that no unifier exists. The terms
are normalized by instantiating the head variable with its binding and then
reducing the potential top-level $\beta$-redex that appears. This instantiation
and reduction process is repeated
until there are no more top-level $\beta$-redexes and the head is
not a variable bound to some term. Then the term with shorter
$\lambda$ prefix is expanded (only on the top-level) so that both have the
same length of $\lambda$ prefix. Finally, the $\lambda$ prefix is ignored, and we
focus only on the body. In this way, we avoid fully substituting
and normalizing terms and perform just enough operations
to determine the next step of the procedure.

If either term of the constraint is flex, we first invoke oracles to solve the
constraint. \ehohii{} implements the most efficient oracles implemented in
Zipperposition:\ fixpoint and pattern oracle (Sect.~\ref{sec:unif:implementation}). An oracle can return three results:
(1)~there is an MGU for the pair (\cn{Unifiable}), which is recorded in
\emph{subst}, and the next pair in $\emph{constraints}$ is tried;
(2)~no MGU exists
for the pair (\cn{NotUnifiable}), which causes the iterator to backtrack;
% to previous state;
(3)~if the pairs do not belong to the subclass that oracle
can solve (\cn{NotInFragment}), we generate possible variable bindings---that is,
we guess the approximate form of the solution.

\ehohii{} has a special module that generates bindings (\cn{NextBinding}). This
module is given the current constraint and the values of \emph{branch\_iter} and
\emph{steps}, and it either returns the next binding and the new values of
\emph{branch\_iter} and \emph{steps} or reports that all different variable
bindings are exhausted. The bindings that \ehohii{}'s unification procedure creates
are imitation, Huet-style projection, identification, and elimination (one
argument at a time) (Sect.~\ref{sec:unif:the-unification-procedure}). A limit on the
total number of applied binding rules can be set, as well as a limit on the
number of individual rule applications. The binding module checks whether limits
are reached using the iterator's \vn{steps} field.

Computing bindings is the only point in the procedure where the search tree
branches and different possibilities are explored. Thus, when \ehohii{} follows
the branch indicated by the binding module, it records the state to which it
needs to return, should the followed branch be backtracked. Therese are values
of $\vn{constraints}, \vn{steps}$, and $\vn{subst}$ before the branch is followed
and the value of the \vn{branch\_iter} that points past the followed branch. The
values of \vn{branch\_iter} are either \cn{BindBegin}, which denotes that no
binding was created, intermediate values that \cn{NextBinding} uses to remember
how far through bindings it is, and \cn{BindEnd}, which indicates that all
bindings are exhausted.

If all bindings are exhausted, the procedure checks whether the pair is
flex--flex and both sides have the same head. If so, the pair is decomposed and
constraints are derived from the arguments of the pair. Otherwise, the iterator
backtracks.
% @PETAR(actually the procedure applies bindings to same-head flex-flex pairs,
% so the text is OK -- we do get "stuck" enumerating those flex-flex pairs. The
% limits help here): Maybe explain here that you don't want a pragmatic prover
% like lambdaE to get stuck in wildly explosive flex-flex pairs? --JB
%
If the constraint is rigid--rigid, for unification to succeed the heads of both
sides of the constraint must be the same. Unification then continues
with new constraints derived from the arguments. Otherwise, the iterator must be
backtracked.

\ourpara{Matching} In Ehoh, the matching algorithm is mostly used inside
simplification rules such as demodulation and subsumption \cite{ss-02-brainiac}.
As these rules must be efficiently performed, using any complex matching
algorithm is not a viable option. Instead, we implemented a matching algorithm
for the pattern class of terms \cite{tn-93-patterns} to complement Ehoh's
$\lambda$-free higher-order matching algorithm \cite[Sect.~4]{section-ehoh}.
A term is a \emph{pattern} if each of its free variables either
has no arguments (as in first-order logic) or is applied to distinct De Bruijn
indices.

To determine which of the two algorithms to call (pattern or $\lambda$-free), we
added a cached property \texttt{HasNonPatternVar}, which is set for terms of the
form $X \,
\overline{s}_n$ where $n>0$ and either there exists some $s_i$ that is not a De Bruijn
index or there exist indices $i < j$ such that $s_i = s_j$ is
a De Bruijn index. This property is propagated to the superterms when they are
perfectly shared. This allows later checks if a term belongs to the pattern
class to be performed in constant time.

% When we need to determine if one term matches the other one we first check if
% any of them has a non-pattern applied variables. If so, $\lambda$-free
% higher-order matching is tried. This algorithm is adjusted to treat $\lambda$ prefixes
% as above in unification algorithm (ensuring that $\beta$-redexes can never occur)
% and that free variables are never bound to terms that have loose bound variables.
% Else if the terms have only pattern variables,
% pattern matching algorithm is tried. Note that in this sense, first-order order
% terms behave as pattern terms and thus pattern matching algorithm is run on
% them. For example, on terms $(X \, \cst{b}, \cst{f} \, \cst{a} \, \cst{b})$
% $\lambda$-free matching algorithm is invoked; on $(\cst{f} \, (\lambda x. \,
% \lambda y.\, X \, x), \cst{f} \, \cst{g})$ pattern matching is tried.

We have modified the $\lambda$-free higher-order matching algorithm to treat
$\lambda$ prefixes as above in unification procedure---by bringing the prefixes
to the same length and ignoring them afterwards. This ensures that the
algorithm will never try to match a free variable to a $\lambda$-abstraction,
making sure that $\beta$-redexes never appear (as in original $\lambda$-free
higher-order matching). We also modified the algorithm to ensure that free variables
are never bound to terms that have loose bound variables. This algorithm cannot
find many complex matching substitutions (matchers), but it can
efficiently determine whether two terms are variable renamings of each other
or whether a simple matcher can be used as in the case of $(X \, (\lambda x. \,x) \,
\cst{b}, \cst{f} \, (\lambda x. \,x) \, \cst{b})$, where ${X \mapsto \cst{f}}$ is
usually the desired matcher. If this algorithm does not find a matcher and both
terms are patterns, pattern matching is tried.

\ourpara{Indexing} Ehoh, like other modern theorem provers, efficiently retrieves
unifiable or matchable pairs of terms using indexing data structures. To find
terms unifiable with a query term or instances of a query term it uses
\emph{fingerprint indexing} \cite{ss-12-fp-indexing}. We extended this data
structure to support (full) higher-order terms in Zipperposition
\cite[Sect.~6]{unif-section}. We used the same approach in \ehohii{},
and we extended feature vector indices \cite{ss-2013-feature-vector}
in the same way.

Ehoh uses \emph{perfect discrimination trees} \cite{mcc-92-pdts} to find
generalizations of the query term (i.e., terms of which the query term is an instance).
This data structure is a trie
that indexes terms by representing them in a serialized, flattened form.
%\begin{rep}, where each node corresponds to a symbol in a term\end{rep}.
The left branch from the root in Figure \ref{fig:pdt} shows how the first-order terms
$\cst{f}\, \cst{a} \, X$ and $\cst{f}\, \cst{a} \, \cst{a}$ are stored.
In Ehoh, this data structure is extended to support partial application
and applied variables \cite{section-ehoh}.

\begin{figure}[tb]
\centering
\begin{tikzpicture}[level distance=3em, sibling distance=7.5em,
  %  edge from parent path={(\tikzparentnode.south) -- ++(0,-0.5em)
  %   -| (\tikzchildnode.north)},
    every node/.style = {shape=rectangle, rounded corners=.75em, minimum size=1.75em, draw, align=center}]
    \node {\phantom{x}}
      child { node {$\cst{f}$}
        child[sibling distance=2.5em] { node %[fill=verylightgray]
          {$\cst{a}$}
          child {node {$X$}}
          child {node {$\cst{a}$}} }
      }
      child { node {$\cst{g}$}
        child {node {$X \, \cst{a} \, \cst{b}$}
               child {node {$\cst{c}$}}}
      }
      child { node {$\cst{h}$}
        child {node {$\internallam_\iota$}
          child {node {$\internallam_\iota$}
            child {node {$X \, \db{0} \, \db{1}$}}
          }
      }}
      ;
\end{tikzpicture}
\caption{First-order, $\lambda$-free higher-order, and higher-order pattern
  terms in a perfect discrimination tree}
\label{fig:pdt}
\end{figure}

In \ehohii{}, we extended this structure to support $\lambda$-abstractions and
the higher-order pattern matching algorithm.
To this end, we changed the way in which terms are
serialized. First, we require that all terms are fully $\eta$-expanded (except
for arguments of variables applied in patterns). Then, when the term is serialized,
we dedicate a separate node for applied variable terms $X \, \overline{s}_n$,
instead of the node for $X$ followed by nodes for serialization of arguments
$\overline{s}_n$. We serialize $\lambda$-abstraction $\lambda x.\, s$ using
a special node $\internallam_\tau$, where $\tau$ is the type of $x$,
followed by the serialization of $s$. Other than these two changes, serialization
remains as in Ehoh, following the gracefulness principle.
Figure \ref{fig:pdt} shows how $\cst{g} \,
(X\,\cst{a}\,\cst{b}) \, \cst{c}$ and $\cst{h} \, (\lambda x. \, \lambda y. \, X
\, y \, x)$ are serialized.

Since the terms are stored in serialized form, it is hard to manipulate
$\lambda$ prefixes of stored terms during matching. Performing $\eta$-expansion
when serializing terms makes sure that matchable terms have
$\lambda$ prefixes of the same length.

We have dedicated separate nodes for applied variables because access to arguments
of applied variable is necessary for the pattern matching algorithm. Even though
arguments can be obtained by querying the arity $n$ of the variable and taking
next $n$ arguments in the serialization, this is both inefficient and inelegant.
As for De Bruijn indices, we treat them the same as function symbols and assign
them their own nodes.

\newcommand{\tstack}{\ensuremath{\texttt{term\_stack}}}
\newcommand{\tproc}{\ensuremath{\texttt{term\_proc}}}

Following the notation from the extension of perfect discrimination trees to
$\lambda$-free higher-order logic \cite{section-ehoh}, we now describe how enumeration of
generalizations is performed. To traverse the tree, \ehohii{} begins at the root
node and maintains two stacks:\ \tstack{} and \tproc{}, where \tstack{} contains the
subterms of the query term that have to be matched, and \tproc{} contains
processed terms that are used to backtrack to previous states. Initially,
\tstack{} contains the query term, the current matching substitution
$\sigma$ is empty, and the successor node is chosen among the child nodes as
follows:

\begin{enumerate}
\item[A.] If the node is labeled with a symbol $\xi$ (where
  $\xi$ is either a De Bruijn index or a constant) and the top item $t$
  of \tstack{} is of the form $\xi \, \overline{t}_n$, replace $t$ by
  $n$~new items $t_1,\dots,t_n$, and push $t$ onto \tproc{}.
\smallskip
\item[B.] If the node is labeled with a symbol $\internallam_\tau$ and the top
  item $t$ of \tstack{} is of the form $\lambda x.\, s$ and the type of $x$ is
  $\tau$, replace $t$ by $s$, and push $t$ onto \tproc{}.
\smallskip
\item[C.] If the node is labeled with a possibly applied variable $X \, \overline{s}_n$
  (where $n \geq 0$), and the top item of \tstack{} is $t$, the 
  matching algorithm described above is run on $X \, \overline{s}_n$ and $t$.
  The algorithm takes into account $\sigma$ built so far and extends it
  if necessary. If the algorithm succeeds, pop $t$ from \tstack{}, push it onto \tproc{},
  and save the original value of $\sigma$ in the node.
\end{enumerate}

Backtracking works in the opposite direction: If the current node is labeled with a De
Bruijn index or function symbol node of arity $n$, pop $n$ terms from \tstack{} and move
top of \tproc{} to \tstack{}. If the node is labeled with
$\internallam_\tau$, pop the top of \tstack{} and move the top of \tproc{} to
\tstack{}. Finally, if the node is labeled with a possibly applied variable, move top of the
\tproc{} to \tstack{} and restore the value of $\sigma$.

As an example of how finding a generalization works, consider the following
states of stacks and substitutions, which emerge when looking for generalizations of 
$\cst{g} \, (\cst{f} \, \cst{a} \, \cst{b}) \, \cst{c}$ in the tree of Figure \ref{fig:pdt}:
%
\[\begin{array}{@{}l@{\kern1em}l@{\kern1em}l@{\kern1em}l@{\kern1em}l@{}}
             & \epsilon                                                & \cst{g}                                                 & \cst{g}.(X \, \cst{a} \, \cst{b})                                                          & \cst{g}.(X \, \cst{a} \, \cst{b}).\cst{c} \\[.5\jot]
  \sigma{:}  & \emptyset                                               & \emptyset                                               & \{X \mapsto \cst{f}\}                                                                      & \{X \mapsto \cst{f}\} \\
  \tstack{:} & [\cst{g} \, (\cst{f} \, \cst{a} \, \cst{b}) \, \cst{c}] & [\cst{f} \, \cst{a} \, \cst{b}, \cst{c}]                & [\cst{c}]                                                                                  & [] \\
  \tproc{:}  & []                                                      & [\cst{g} \, (\cst{f} \, \cst{a} \, \cst{b}) \, \cst{c}] & [\cst{f} \, \cst{a} \, \cst{b}{,}\; \cst{g} \, (\cst{f} \, \cst{a} \, \cst{b}) \, \cst{c}] & [\cst{c}{,}\;\cst{f} \, \cst{a} \, \cst{b}{,}\; \cst{g} \, (\cst{f} \, \cst{a} \, \cst{b}) \, \cst{c}]
  \end{array}\]
